---
title: "Untitled"
format: html
editor: visual
---

## Setup

```{r}
library(tidyverse)
healthdata <- readr::read_csv("data/nhanes.csv")
```

# Mappings

In `ggplot2`, the mapping specifies how variables (columns) are mapped onto aesthetic attributes. The mapping between variables and aesthetics is accomplished with the `aes` function. Example:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point()
```

## Adding aesthetics

Besides x and y, you may want to map variables to other aesthetic attributes: color, fill, shape, line type, size, line width, label, etc. This is useful, for example, to show how a relationship between two numeric variables varies by a categorical variable.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight, color = gender)) +
  geom_point()
```

You can even have two categorical variables at the same time, although it can quickly become too much for one plot.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight, color = gender, shape = physically_active)) +
  geom_point()
```

### TRY IT

Create a scatter plot (`geom_point`) of `cholesterol` (`y`) against `weight` (`x`).

```{r}

```

Map `gender` to the `color` aesthetic.

```{r}

```

## Different aesthetics for different `geom`s

You can use different aesthetics for different geoms. This is useful, for example, when you want to map certain aesthetics (like fill or color) only to certain layers of your plot rather than to all geoms.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(aes(fill = physically_active), bins = 15)
```

# Geoms

After mapping data to a plot (with `aes`), you need to display it in a graph. This involves determining how the data are displayed (with points, lines), whether you need to compute new variables from the data (statistical transformations), and where you want to display pieces of data. For the most part, you'll specify this using functions that start with `geom_`, followed by the type of plot.

This notebook mostly presents various common `geom`s and for what type of variables they are appropriate. Towards the end we'll also see that we can use more than one `geom` and that we can change the styling within `geom`.

## Continuous data

This section presents `geom`s that are appropriate for continuous data.

### One variable

This section presents `geom`s that are appropriate to show the distribution of a single continuous variable. Using these is a good practice to understand your data before you go into more complex forms of analysis.

#### Histogram

Histograms are probably the more common way to plot the distribution of a continuous variable. For example, they are a great plot to explore the distribution of the outcome variable in a regression model.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram()
```

By default, `geom_histogram` uses 30 bins. However, most times you'll want to decide a better number for your data. You'll want to select a number that shows the shape of your data clearly, without over-emphasizing noise but without missing important patterns. You can do some exploration to decide.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(bins = 100)
```

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(bins = 5)
```

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(bins = 20)
```

#### Density plot

A density plot is a smoothed version of a histogram using kernel density estimates. It can be nice if you want to see the overall pattern of your data.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_density()
```

### Two variables

This section presents common `geom`s to display the relation between two continuous variables.

#### Scatter plot

A scatter plot is probably the most common way to display the relationship between two continuous variables. For example, it is a great tool for exploratory data analysis as you try to see which variables seem to be related.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point()
```

If you have too much data, you may have an issue of overplotting -- or having the points overlap too much and not being able to see the relationship. You can explore the argument `alpha` in `geom_point` and `geom_jitter`.

#### Line plot

Line plots show a trend across values of a variable - most commonly across time, but we don't have a time variable in this data. We'll compute some values across ages instead first so that we have appropriate data for a line plot:

```{r}
by_age <- healthdata %>%
  group_by(age) %>%
  summarize(avg_pulse = mean(pulse, na.rm=TRUE))

by_age
```

```{r}
ggplot(data = by_age, aes(x = age, y = avg_pulse)) +
  geom_line()
```

## Discrete data

This section presents common `geom`s to display discrete data.

### One variable

#### Bar Plot

Bar plots are by far the most common way to visualize a single discrete variable. In a way, you can think of bar plots like the histograms of discrete data. However, the "bins are pre-defined" in this case, because it's about counting the number of observations for each category. It's basically a visual display of a table.

```{r}
ggplot(data = healthdata, aes(x = physically_active)) +
  geom_bar()
```

## Continuous and discrete data

This section presents common plots to visualize the relationship between continuous and discrete variables.

### One discrete, one continuous

This section presents common plots to visualize the relationship between one discrete and one continuous variables.

#### Side by side box plot

Box plots compactly display the distribution of a continuous variable. They visualize the median, the 25th and 75th percentiles, and outliers (lower or higher than 1.5 times the interquartile range, i.e., the range between the 25th and the 75th percentiles). Side by side box plots display the distribution of a continuous variable by a discrete variable. They are a great way to explore the relationship between continuous and discrete variables as part of the exploratory data analysis or potentially even to present, say, experimental results.

```{r}
ggplot(data = healthdata, aes(x = physically_active, y = weight)) +
  geom_boxplot()
```

#### Side by side violin plot

A violin plot is a compact display of a continuous distribution. They are a mix of a box plot and a density plot. Compared to side by side box plots, side by side violin plots are nice if you want to see the shape of the distribution of the continuous variable.

```{r}
ggplot(data = healthdata, aes(x = physically_active, y = weight)) +
  geom_violin()
```

#### Histogram by group

Histograms are a great way to explore the distribution of a continuous variable. You can also add a discrete variable, either stacking the bars on top of each other or overlapping them.

Let's focus on stacked histograms. In this case, the height of the bar represents the total count and the color represents how that total is distributed across the discrete variable.

```{r}
ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(aes(fill = physically_active), bins = 20)
```

### One discrete, two continuous

This section presents common `geom`s to visualize the relationship between one discrete and two continuous variables.

#### Scatter plot by group

Remember that a scatter plot is a great way to visualize the relationship between two continuous variables. You can add a third, discrete variable as the color or shape aesthetic to see how the relationship between the two continuous variables varies by the discrete variable.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight, color = physically_active)) +
  geom_point()
```

#### Line plot by group

Line plots show a trend across values of a variable - most commonly across time, but we don't have a time variable in this data. Earlier we computed some values across ages to have appropriate data for a line plot. Now we're computing across ages and whether the person is physically active.

```{r}
by_age_physically_active <- healthdata %>%
  group_by(age, physically_active) %>%
  summarize(avg_pulse = mean(pulse, na.rm=TRUE))

by_age_physically_active
```

With these data, we can now break the relationship between average pulse and age by whether the person is physically active.

```{r}
ggplot(data = by_age_physically_active, aes(x = age, y = avg_pulse, color = physically_active)) +
  geom_line()
```

Be careful when adding too many groups, because it can quickly become too crowded/hard to read. You don't want to make people have to spend too much time trying to understand what your plot is meant to convey.

## More than one `geom`

You can use multiple `geom`s. This is useful when you want to display different types of information on the same plot, such as raw data points with a trend line.

Let's start with a scatter plot:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point() 
```

There seems to be a relation. It would be nice to have a visual way to summarize it. We can add another `geom` to the plot, specifically `geom_smooth`.

`geom_smooth` adds a line to your plot that helps you see the general trend in your data. Although we won't go into all the details here, it works by fitting a model to your data and plotting the predicted values. Depending on how much data you have and whether you change the defaults, there are different models that can be fitted to the data.

By default, `geom_smooth` produces a smoothed line with a confidence band:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point() +
  geom_smooth()
```

You can change the smoothing method used with the `method` argument. For example, you can use a linear regression:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm")
```

You can also remove the confidence band with `se = FALSE`:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

### TRY IT

Take a look at the variables available in `healthdata`.

```{r}
names(healthdata)
# View(healthdata)
```

Think of a relationship that would be interesting to explore and, based on the type of variables, create a plot with one or more of the `geom`s presented in this notebook.

```{r}

```

## Changing the styling within `geom`s

Scale functions, `scale_<aesthetic>_<type>`, control visual elements that are about how data are represented. For example, what color palette to use for a discrete variable.

The theme, `theme_*` and `theme` functions, controls the style of plot elements that do not represent data (backgrounds, axes, grids, etc.).

But what if you just want to change the colors of all of the points in a scatter plot - for example, use purple instead of black dots?

Within `geom_*` you can control the style of elements that affect a specific `geom` and that are independent of the data.

### One `geom`

#### Color

For example, you can change the color of the points in a scatter plot to adhere to a style guide.

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point(color = "purple") 
```

#### Shape

You can also change the [shape of the points](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#sec:shape-spec):

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point(shape = 4) 
```

#### Size

You can also change the size:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point(size = 5) 
```

#### Transparency

When you have overplotting, you can change the transparency with `alpha` -- that way, you can see that darker regions have more points and lighter regions have less points:

```{r}
ggplot(data = healthdata, aes(x = height, y = weight)) +
  geom_point(size = 5, alpha = 0.5) 
```

### More than one `geom`

If you have more than one `geom`, you can change the styling for each `geom` independently. The styling options depend on each `geom`. For example, `geom_point` doesn't have `linetype`, while `geom_line` does.

Let's look at an example with `geom_line` and `geom_point`:

```{r}
ggplot(data = by_age, aes(x = age, y = avg_pulse)) +
  geom_line() +
  geom_point()
```

Let's add some of what we already know for points:

```{r}
ggplot(data = by_age, aes(x = age, y = avg_pulse)) +
  geom_line() +
  geom_point(color = "green", shape = 8, size = 4)
```

#### Line type

You can change the [line type](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#sec:line-type-spec):

```{r}
ggplot(data = by_age, aes(x = age, y = avg_pulse)) +
  geom_line(linetype = "dashed") +
  geom_point(color = "green", shape = 8, size = 4)
```

#### Line width

You can also change the line width:

```{r}
ggplot(data = by_age, aes(x = age, y = avg_pulse)) +
  geom_line(linetype = "dashed", linewidth = 2) +
  geom_point(color = "green", shape = 8, size = 4)
```

### Wrap-up for styling

Hopefully it's clear now that within `geom_*` you can control the style of elements that affect a specific `geom` and that are independent of the data. What you can control depends on the specific `geom` -- what makes sense for that type of plot. And you control the styling for each `geom` independently.
