---
title: "Untitled"
format: html
editor: visual
---

## Setup

```{r}
library(tidyverse)
healthdata <- readr::read_csv("data/nhanes.csv")
```

## Combining Multiple Plots

`ggplot2` has the ability to plot different subsets of the data using the same visualization settings (with `facet_`). However, what if you have two separate plots (even from different data) and you want to create a figure for a paper that combines both plots? This notebooks shows how to do that.

## `patchwork`

[`patchwork`](https://patchwork.data-imaginist.com/) is a package that makes it easy to combine separate `ggplot`s into the same figure.

If you don't have `patchwork` already installed, you need to install it first:

```{r, eval=FALSE}
install.packages("patchwork")
```

Let's load `patchwork` to be able to use it:

```{r}
library(patchwork)
```

Let's create two `ggplot`s that we will combine into one figure. We're going to save the plots to two variables (`p1` and `p2`) so that we can use those variables later when we combine the plots.

```{r}
p1 <- ggplot(data = healthdata, aes(x = weight)) +
  geom_histogram(bins = 20) +
  labs(title = "Plot 1")

p1
```

```{r}
p2 <- ggplot(data = healthdata, aes(x = height, y = weight, color = physically_active)) +
  geom_point() +
  labs(title = "Plot 2")

p2
```

Combine the two plots into one, arranging them horizontally into a row:

```{r}
p1 + p2
```

Combine the two plots into one, arranging them vertically in a column:

```{r}
p1 / p2
```

### Overall Title and Caption

Let's add an overall title and caption to the figure:

```{r}
p1 + p2 +
  plot_annotation(title = "Two cool plots",
                  caption = "NHANES data")
```

### Changing legend position

Let's move the legend from the right to the bottom. Since this is about visual aspects of the plot that don't depend on the data, we can use `theme`, specifically `legend.position`:

```{r}
p1 + p2 +
  plot_annotation(title = "Two cool plots",
                  caption = "NHANES data") +
  theme(legend.position = "bottom")
```

## TRY IT

Use `patchwork` to put the two plots below together. Add a title.

```{r}
pe1 <- ggplot(data = healthdata, aes(x = marital_status, y = age)) +
  geom_boxplot()

pe2 <- ggplot(data = healthdata, aes(x = marital_status)) +
  geom_bar()

```
